Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-01 _tex/index.tex
Built site for gh-pages

{In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher level of the trait. After conducting multiple rounds
of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. CJ has
been successfully employed in assessing the quality of written texts
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits being measured
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aim to address all three concerns simultaneously.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The literature indicates that these scores have been used to identify
`misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the latent trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting additional
analyses and tests can inflate their precision and statistical power,
increasing the risk of committing a type I error \citep{McElreath_2020},
which is when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
recommend conducting these analyses and tests within a structural model
that accounts for both the scores and their uncertainties, rather than
treating them separately. Thus, an integrated approach combining CJ's
structural and measurement models can offer significant advantages.

>>>>

{In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher trait level. After conducting multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Additionally, research on validity suggests that scores generated by CJ
can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023},
while research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aim to address all three concerns simultaneously.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The literature indicates that the scores have been used to identify
`misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power, increasing
the risk of committing a type I error \citep{McElreath_2020}, which is
when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
recommend conducting these analyses and tests within a structural model.
A structural model specify how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
structural and measurement models can offer significant advantages.
